---
title: "Statistical Learning Assignment 2"
author: "Michael J Jones"
date: "24/04/2015"
output: pdf_document
---

# Question 1
Define a Cox Proportional Hazard Model (M1) for the covariates: clinic, prison, dose.

***

```{r}
library(survival)

load("addicts.rda")

M1 <- coxph(formula = Surv(time = survt, event = status) ~ clinic + prison + dose,
            data = dat)

summary(M1)
```

# Question 2
Perform a regression analysis for the model M1 and provide a discussion of the results. Remark: Follow the 
instructions given in Tutorial 8.

***

<!-- TODO: Find / Ask for Tutorial 8 -->

# Question 3

## Part A
Check the proportional hazard assumption of M1 and adjust the model if necessary

***

There are at least two methods for making sure that our model (M1) conforms to the *proportional hazard assumption*.
The proportional hazard assumption...etc...discuss simple and hard way

### Simple way
```{r}
library(GGally)

ggsurv(survfit(formula = Surv(time = survt, event = status) ~ clinic,
            data = dat))

ggsurv(survfit(formula = Surv(time = survt, event = status) ~ prison,
            data = dat))

ggsurv(survfit(formula = Surv(time = survt, event = status) ~ dose,
            data = dat))
```
